Supplemental Information

### 2022-02-17 ms/manuscript.Rmd
Add SI fig with Du et al 2021 dates

tar_load(
	c(
		accs_exclude,
		fern_sanger_seqs_raw,
		raw_fasta_all,
		sanger_seqs_combined_filtered,
		sanger_seqs_rogues_removed,
		sanger_accessions_selection,
		plastome_seqs_combined_filtered,
		plastome_metadata_raw_all,
		plastome_metadata_renamed,
		sanger_alignment,
		plastome_alignment,
		plastome_metadata_renamed,
		ppgi_taxonomy,
		plastid_calibration_dates,
		pteridocat,
		sanger_alignment_tbl,
		plastome_alignment_tbl,
		monophy_by_clade,
		plastome_tree,
		plastid_tree_dated,
		fossil_calibration_tips,
		treepl_cv_results),
	store = ftol_cache
)

@@

We dated the ML tree using treePL [@Smith2012].
We selected `r n_calibrations_ferns` fern fossils to use as calibration points from a curated list of `r n_fern_fossils_total` fern fossil taxa.
In the case of redundant fossils (those calibrating the same node in the phylogeny; `r eg`, stems of families that are sister to each other, such as Dipteridaceae and Matoniaceae) we selected the fossil with the oldest age (`r ie`, the upper limit of the oldest stratigraphic period assigned to the fossil). 
We assigned fossils to lineages only after consulting the original publication (as opposed to only using the fossil species or genus) so we could reassess the identification relative to changes in taxonomy and hypotheses of phylogenetic relationships.
Taxon concepts may vary between the original publication and currently applied taxonomy (`r eg`, Dicksoniaceae *sensu lato* including other tree fern families used in description of the fossil but Dicksoniaceae *sensu stricto* used currently) and hypotheses of phylogenetic relationships among extant species may change (`r eg`, *Dennstaedtia* used in description of a fossil but this genus now known to be polyphyletic); in both cases, the resulting node to be calibrated for a given fossil will also change.
We also applied two calibration points outside of ferns including the most recent common ancestor (MRCA) of euphyllophytes (`r fossils$Euphyllophytes$age` Ma) and MRCA of tracheophytes (`r fossils$Tracheophytes$age` Ma), and fixed the root age (land plants) at 475 Ma [@Testo2016a].
All calibration points other than the root are minimum ages.

@@

We recovered older crown ages for ferns and other large clades as well as older stem ages for families relative to previous studies (`r figure("divtimes")`).
This may be due to the fact that we used a completely revised set of fossil calibration points relative to previous studies, which not only resulted in a more densely calibrated tree but also a higher number of fern families with minimum fossil ages.
Several of the recent studies analyzing divergence times for global fern phylogenies all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
However, we found several of these calibration points were not appropriately applied due to differences in taxonomic concepts between the original fossil publication and the phylogeny in which they were used.
Moreover, our set of calibration points more than doubles the number used by previous studies.
It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using the same (or very similar), more limited set of calibration points (SI Fig 1).

<<FIXME: add discussion of specific groups once we have final ages>>

@@

```{r plot-prep}
# Make vector of outgroup taxa
og_taxa <-
	sanger_sampling %>%
	filter(outgroup == TRUE) %>%
	pull(species)

@@

clade_cols_tib <-
	tibble(
		color = clade_cols,
		major_clade = names(clade_cols)
	) %>%
	mutate(
		code = substr(major_clade, 1, 2),
		code = case_when(
			major_clade == "Saccolomatineae" ~ "Sac",
			major_clade == "Salviniales" ~ "Sal",
			TRUE ~ code
		)) %>%
	assert(is_uniq, code)
```

@@

# Tree plot ----
# Make tibble for annotating major clades
clade_tibble <- 
	# Use assess_monophy to get MRCA of each major clade
	# They should all by monophyletic anyways
	select(sanger_sampling, species, major_clade) %>%
	assess_monophy(
		taxon_sampling = .,
		tree = plastid_tree_dated,
		tax_levels = "major_clade"
	) %>%
	get_result_monophy(1) %>%
	filter(!taxon %in% c(
		"Selaginellales", "Lycopodiales", "Isoëtales", "Isoetales")) %>%
	verify(all(monophyly == "Yes")) %>%
	transmute(major_clade = taxon, mrca = parse_number(mrca)) %>%
	left_join(clade_cols_tib, by = "major_clade")

@@

```{r div-times-make}
# Convert family ages to long format
family_ages_median <-
family_stem_ages %>%
	select(family, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
		left_join(
		unique(select(sanger_sampling, family, major_clade)),
		by = "family"
	) %>%
	left_join(clade_cols_tib, by = "major_clade"
	) %>%
	mutate(family = fct_reorder(family, current_best)) %>%
	pivot_longer(values_to = "age", -c(family, color, code, major_clade)) %>%
	separate(name, into = c("author", "age_type"), sep = "_") %>%
	assert(not_na, author, age_type) %>%
	filter(age_type %in% c("best", "median")) %>%
	assert(not_na, major_clade, color)

# Convert family errors to almost long format
family_ages_errors <-
	family_stem_ages %>%
	select(family, current_best = age) %>%
	left_join(other_dates, by = "family") %>%
	mutate(family = fct_reorder(family, current_best)) %>%
	select(family, matches("ts|rothfels")) %>%
	select(-matches("median"))

family_ages_errors <-
	bind_rows(
		transmute(family_ages_errors, family,
					 high = ts_high, low = ts_low, author = "ts"),
		transmute(family_ages_errors, family,
					 high = rothfels_high, low = rothfels_low, author = "rothfels")
	)

@@

# Format y-axis text: show major clade in parentheses
family_labs_tbl <-
family_ages_median %>%
	select(family, code) %>%
	unique() %>%
	mutate(label = glue("{family} ({code})"))
family_labs <- family_labs_tbl$label %>%
	set_names(., family_labs_tbl$family)

@@

div_times_plot <-
ggplot(mapping = aes(y = family)) +
	geom_errorbarh(
		data = family_ages_errors,
		aes(xmax = high, xmin = low, color = author),
		position = position_dodge(width = 0.5)
	) +
	geom_point(
		data = filter(family_ages_median, age_type %in% c("best", "median")),
		aes(x = age, color = author, shape = author),
		position = position_dodge(width = 0.5)) +
	scale_shape_manual(
		values = ages_shapes,
		limits = ages_sampling$source, 
		labels = ages_sampling$label) +
	scale_color_manual(
		values = ages_colors,
		limits = ages_sampling$source, 
		labels = ages_sampling$label) +
	scale_y_discrete(labels = family_labs) +
	scale_x_reverse("Age (millions of years)") +
	coord_geo(
		dat = list("periods", "eras"),
		pos = list("b", "b"),
		abbrv = list(TRUE, FALSE),
		xlim = c(500, 0),
		fill = "transparent",
		color = "grey20",
		lab_color = "grey30",
		height = unit(1, "lines"),
		size = 3) +
	guides(shape = guide_legend(nrow = 2)) +
	theme_bw() +
	theme(
		axis.title.y = element_blank(),
		legend.title = element_blank(),
		legend.position = "bottom"
	)
```

>>>>

tar_load(
	c(
		accs_exclude,
		fern_sanger_seqs_raw,
		raw_fasta_all,
		sanger_seqs_combined_filtered,
		sanger_seqs_rogues_removed,
		sanger_accessions_selection,
		plastome_seqs_combined_filtered,
		plastome_metadata_raw_all,
		plastome_metadata_renamed,
		sanger_alignment,
		plastome_alignment,
		plastome_metadata_renamed,
		ppgi_taxonomy,
		plastid_calibration_dates,
		pteridocat,
		sanger_alignment_tbl,
		plastome_alignment_tbl,
		monophy_by_clade,
		plastome_tree,
		plastid_tree_dated,
		fossil_calibration_tips,
		ts_fossil_calibration_tips,
		treepl_cv_results),
	store = ftol_cache
)

@@

We rooted the ML tree on bryophytes and estimated divergence dates using penalized likelihood with treePL [@Smith2012].
We selected `r n_calibrations_ferns` fern fossils to use as calibration points from a curated list of `r n_fern_fossils_total` fern fossil taxa.
In the case of redundant fossils (those calibrating the same node in the phylogeny; `r eg`, stems of families that are sister to each other, such as Dipteridaceae and Matoniaceae) we selected the fossil with the oldest age (`r ie`, the upper limit of the oldest stratigraphic period assigned to the fossil). 
We assigned fossils to lineages only after consulting the original publication (as opposed to only using the fossil species or genus) so we could reassess the identification relative to changes in taxonomy and hypotheses of phylogenetic relationships.
Taxon concepts may vary between the original publication and currently applied taxonomy (`r eg`, Dicksoniaceae *sensu lato* including other tree fern families used in description of the fossil but Dicksoniaceae *sensu stricto* used currently) and hypotheses of phylogenetic relationships among extant species may change (`r eg`, *Dennstaedtia* used in description of a fossil but this genus now known to be polyphyletic); in both cases, the resulting node to be calibrated for a given fossil will also change.
We also applied two calibration points outside of ferns including the most recent common ancestor (MRCA) of euphyllophytes (`r fossils$Euphyllophytes$age` Ma) and MRCA of tracheophytes (`r fossils$Tracheophytes$age` Ma), and fixed the root age (land plants) at 475 Ma [@Testo2016a].
All calibration points other than the root are minimum ages.

@@

We recovered older crown ages for ferns and other large clades as well as older stem ages for families relative to previous studies (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).
This may be due to the fact that we used a completely revised set of fossil calibration points relative to previous studies, which not only resulted in a more densely calibrated tree but also a higher number of fern families with minimum fossil ages.
Several of the recent studies analyzing divergence times for global fern phylogenies all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
However, we found several of these calibration points were not appropriately applied due to differences in taxonomic concepts between the original fossil publication and the phylogeny in which they were used.
Moreover, our set of calibration points more than doubles the number used by previous studies.
It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using the same (or very similar), more limited set of calibration points (Figures `r figure_num("divtimes")`, `r s_figure_num("du-ages")`).

<<FIXME: add discussion of some specific groups once we have final ages>>

@@

```{r plot-prep}
## ---- plot-prep-main

# Make vector of outgroup taxa
og_taxa <-
	sanger_sampling %>%
	filter(outgroup == TRUE) %>%
	pull(species)

@@

clade_cols_tib <-
	tibble(
		color = clade_cols,
		major_clade = names(clade_cols)
	) %>%
	mutate(
		code = substr(major_clade, 1, 2),
		code = case_when(
			major_clade == "Saccolomatineae" ~ "Sac",
			major_clade == "Salviniales" ~ "Sal",
			TRUE ~ code
		)) %>%
	assert(is_uniq, code)

# Make tibble for annotating major clades
clade_tibble <- 
	# Use assess_monophy to get MRCA of each major clade
	# They should all by monophyletic anyways
	select(sanger_sampling, species, major_clade) %>%
	assess_monophy(
		taxon_sampling = .,
		tree = plastid_tree_dated,
		tax_levels = "major_clade"
	) %>%
	get_result_monophy(1) %>%
	# Exclude lycophytes
	filter(!taxon %in% c(
		"Selaginellales", "Lycopodiales", "Isoëtales", "Isoetales")) %>%
	verify(all(monophyly == "Yes")) %>%
	transmute(major_clade = taxon, mrca = parse_number(mrca)) %>%
	left_join(clade_cols_tib, by = "major_clade")
```

@@

# Tree plot ----

@@

```{r div-times-make}
# Convert family ages to almost-long format

# Helper function for subsetting dates to get into
# almost-long format
subset_ages <- function(other_dates, author_sel) {
	other_dates %>%
		select(family, contains(glue("{author_sel}_"))) %>%
		rename_with(~str_remove_all(., glue("{author_sel}_"))) %>%
		mutate(author = author_sel)
}

# Combine dates in almost-long format:
# one row per family/study combination,
# cols for median, high, and low date
all_ages <-
bind_rows(
	subset_ages(other_dates, "ts"),
	subset_ages(other_dates, "rothfels"),
	select(other_dates, family, median = schuettpelz_best) %>%
		mutate(author = "schuettpelz"),
	transmute(
		family_stem_ages,
		family,
		author = "current",
		median = age
	)
) %>%
	left_join(
		unique(select(sanger_sampling, family, major_clade)), by = "family") %>%
	left_join(clade_tibble, by = "major_clade") %>%
	# (some median dates are NA if that family wasn't included in a study)
	assert(not_na, family, major_clade, code, author) %>%
	# Set family as a factor ordered by current age estimate
	left_join(family_stem_ages, by = "family") %>%
	mutate(family = fct_reorder(family, age)) %>%
	select(-age)

@@

# Format y-axis text: show major clade in parentheses
family_labs_tbl <-
all_ages %>%
	select(family, code) %>%
	unique() %>%
	mutate(label = glue("{family} ({code})"))

family_labs <- family_labs_tbl$label %>%
	set_names(., family_labs_tbl$family)

@@

div_times_plot <-
ggplot(all_ages) +
	geom_pointrange(
		aes(
			x = median, xmin = low, xmax = high,
			y = family,
			color = author, shape = author),
		position = position_dodge(width = 0.5)) +
	scale_shape_manual(
		values = ages_shapes,
		limits = ages_sampling$source, 
		labels = ages_sampling$label) +
	scale_color_manual(
		values = ages_colors,
		limits = ages_sampling$source, 
		labels = ages_sampling$label) +
	scale_y_discrete(labels = family_labs, expand = expansion(add = 0.7)) +
	scale_x_reverse("Age (millions of years)") +
	coord_geo(
		dat = list("periods", "eras"),
		pos = list("b", "b"),
		abbrv = list(TRUE, FALSE),
		xlim = c(500, 0),
		fill = "transparent",
		color = "grey20",
		lab_color = "grey30",
		height = unit(1, "lines"),
		# FIXME: change this to true when issue gets resolved
		# https://github.com/willgearty/deeptime/issues/41
		expand = FALSE,
		size = 3) +
	guides(shape = guide_legend(nrow = 2)) +
	theme_bw() +
	theme(
		axis.title.y = element_blank(),
		legend.title = element_blank(),
		legend.position = "bottom"
	)
```

### 2022-02-17 ms/si.Rmd
Add SI fig with Du et al 2021 dates

tar_load(
	c(
		sanger_alignment_tbl,
		sanger_sampling,
		sanger_accessions_selection),
	store = ftol_cache
)

# Targets from MS workflow
tar_load(
	c(
		gb_species_by_year,
		ref_files,
		other_dates
	),
	store = here::here("_targets")
)
```

@@

@@

>>>>

tar_load(
	c(
		sanger_alignment_tbl,
		sanger_sampling,
		sanger_accessions_selection,
		plastome_tree,
		plastid_tree_dated),
	store = ftol_cache
)

# Targets from MS workflow
tar_load(
	c(
		gb_species_by_year,
		ref_files,
		other_dates,
		du_dates_all,
		family_stem_ages
	),
	store = here::here("_targets")
)
```

@@

```{r plot-prep-load}
# Load the plot prep code from the main MS
knitr::read_chunk(
	here::here("ms/manuscript.Rmd"),
	labels = "plot-prep-main",
	from = "og_taxa <-",
	to = "left_join\\(clade_cols_tib"
	)
```

<!-- Run the plot prep code -->
```{r, plot-prep-main}
```

@@

```{r du-ages, fig.width = 7, fig.height = 5}
# Helper function for subsetting dates to get into
# almost-long format
subset_ages <- function(other_dates, author_sel) {
	other_dates %>%
		select(family, contains(glue("{author_sel}_"))) %>%
		rename_with(~str_remove_all(., glue("{author_sel}_"))) %>%
		mutate(author = author_sel)
}

# Filter Du ages to just stem families
du_dates_stem_fam <-
	du_dates_all %>%
	filter(
		str_detect(affinities, "eae$"),
		affinities_group == "stem",
		str_detect(author, "^r")) %>%
	select(family = affinities, author, median, low, high)

# Make tibble of Du
du_schemes <- du_dates_stem_fam %>%
	select(scheme = author) %>%
	unique() %>%
	mutate(
		code = str_split(scheme, "_") %>%
			map_chr(last) %>%
			str_to_upper()) %>%
	arrange(code)

# Combine dates in almost-long format:
# one row per family/author combination,
# cols for median, high, and low date
all_ages_with_du <-
	bind_rows(
		du_dates_stem_fam,
		subset_ages(other_dates, "ts"),
		subset_ages(other_dates, "rothfels"),
		select(other_dates, family, median = schuettpelz_best) %>%
			mutate(author = "schuettpelz"),
		transmute(
			family_stem_ages,
			family,
			author = "current",
			median = age
		)
	) %>%
	filter(family %in% du_dates_stem_fam$family) %>%
	left_join(
		unique(select(sanger_sampling, family, major_clade)), by = "family") %>%
	left_join(clade_tibble, by = "major_clade") %>%
	# (some median dates are NA if that family wasn't included in a author)
	assert(not_na, family, major_clade, code, author) %>%
	# Set family as a factor ordered by current age estimate
	left_join(family_stem_ages, by = "family") %>%
	mutate(family = fct_reorder(family, age)) %>%
	select(-age)

# Define shapes and colors for points
du_ages_colors <- c(
	current = palette_okabe_ito(order = "vermillion"),
	rothfels = palette_okabe_ito(order = "bluishgreen"),
	schuettpelz = palette_okabe_ito(order = "blue"),
	ts = palette_okabe_ito(order = "orange"),
	rep(
		palette_okabe_ito(order = "gray"),
		length(du_schemes$scheme)) %>%
		set_names(du_schemes$scheme)
)

du_ages_shapes <- c(
	current = 16, # circle
	rothfels = 16,
	schuettpelz = 16,
	ts = 16,
	seq_along(du_schemes$scheme) %>%
		subtract(1) %>%
		set_names(du_schemes$scheme)
)

# Cross doesn't work because we can't distinguish error bars
du_ages_shapes[du_ages_shapes == 3] <- 6

# Format names of papers for printing
du_ages_sampling <-
	tribble(
		~source, ~year, ~label,
		"schuettpelz", 2009, "Schuettpelz & Pryer 2009",
		"rothfels", 2015, "Rothfels et al. 2015",
		"ts", 2016, "Testo and Sundue 2016",
		"current", 2021, "This study",
	) %>%
	bind_rows(
		du_schemes %>%
			mutate(
				year = 2020 + 0.1*seq_along(du_schemes$scheme),
				label = paste("Du et al. 2021 scheme", code)
			) %>%
			transmute(source = scheme, year, label)
	) %>%
	mutate(source = fct_reorder(source, year))

# Format y-axis text: show major clade in parentheses
family_labs_tbl <-
	all_ages_with_du %>%
	select(family, code) %>%
	unique() %>%
	mutate(label = glue("{family} ({code})"))

family_labs <- family_labs_tbl$label %>%
	set_names(., family_labs_tbl$family)

ggplot(all_ages_with_du) +
	geom_pointrange(
		aes(
			x = median, xmin = low, xmax = high,
			y = family,
			color = author, shape = author),
		position = position_dodge(width = 0.5)) +
	scale_shape_manual(
		values = du_ages_shapes,
		limits = du_ages_sampling$source, 
		labels = du_ages_sampling$label) +
	scale_color_manual(
		values = du_ages_colors,
		limits = du_ages_sampling$source, 
		labels = du_ages_sampling$label) +
	scale_y_discrete(labels = family_labs, expand = expansion(add = 0.7)) +
	scale_x_reverse("Age (millions of years)") +
	coord_geo(
		dat = list("periods", "eras"),
		pos = list("b", "b"),
		abbrv = list(TRUE, FALSE),
		xlim = c(250, 0),
		fill = "transparent",
		color = "grey20",
		lab_color = "grey30",
		height = unit(1, "lines"),
		# FIXME: change this to true when issue gets resolved
		# https://github.com/willgearty/deeptime/issues/41
		expand = FALSE,
		size = 3) +
	theme_bw() +
	theme(
		axis.title.y = element_blank(),
		legend.title = element_blank()
	)
```

**`r s_figure("du-ages")`**.
Stem age of fern families (Ma) estimated by selected studies, only including families that were sampled across all studies.
For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a; @Du2021].
For other studies, point indicates best (most likely) estimate [@Schuettpelz2007b\; this study].
@Du2021 used six different dating schemes; for details, see that paper.
Codes in parentheses after family names indicate major clade as in `r figure("ftol")`.
Period name abbreviations as follows: FIXME.
